First Order Alternation

### 2018-07-17 quantifiers/draft.tex
ok

In the upcoming developments it is more convenient to work with
logical formulae defining executions. Let $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the rest of
this session. For any $i \in \nat$, we denote by $\stamp{Q}{i} =
\set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} = \set{\stamp{x}{i}
  \mid x \in X}$ the sets of time-stamped predicates and variables,
respectively. For a formula $\psi$ and $i \in \nat$, we define
$\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
formula in which all input variables and state predicates (and only
those symbols) are replaced by their time-stamped counterparts. Given
a sequence of input events $\alpha = a_1 \ldots a_n$, the \emph{path
  formula} of $\alpha$ (in $\A$) is defined as follows:
\[\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
\bigwedge_{i=1}^n \bigwedge_{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{}
  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \Longrightarrow \stamp{\psi}{i}\]
The automaton $\A$, to which $\pathform{\alpha}$ refers, will always
be clear from the context. To formalize the relation between the
low-level configuration-based execution semantics and the symbolic
path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any
execution forest $\F$ of $\A$ over $w$ is associated an interpretation
$\I_{\F}$ of the set of time-stamped predicates $\set{\stamp{q}{i}
  \mid q \in Q, i \in [n]}$, defined as follows:
\[\I_{\F}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}}
  \mid q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in
    $\F$}}\enspace.\]

@@

  ``$\supseteq$'' Let $\F$ be an execution of $\A$ over $w$. We show
  that $\I_{\F}$ is a minimal interpretation such that $\I_{\F},
  w_\Data \models \pathform{w_\Sigma}$, by induction on $n \geq
  0$. For $n=0$, $\F$ is a cube from $\cube{\minsem{\iota}}$, by
  definition. Then $\I_{\F} \models \stamp{\iota}{0}$ and moreover, it
  is a minimal such interpretation. For the inductive case $n > 0$,
  let $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\G$ be the
  restriction of $\F$ to $u$. Consequently, $\I_\G$ is the restriction
  of $\I_\F$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$. By the inductive
  hypothesis, $\I_{\G}$ is a minimal interpretation such that
  \(\I_{\G}, u_{\Data} \models \pathform{u_\Sigma}\). Since
  $\I_{\F}(\stamp{q}{n}) = \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
    q(d_1,\ldots,d_{\#(q)}) \text{ occurs on the $n$-th level in
      $\F$}}$, we have $\I_{\F}, w_\Data \models \varphi$, for each
  subformula $\varphi = \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$ of
  $\pathform{w_\Sigma}$, by the execution semantics of $\A$. This is
  the case because the children of each node labeled with
  $q(d_1,\ldots,d_{\#(q)})$ on the $(n-1)$-th level of $\F$ form a
  cube from $\cube{\minsem{\psi}_\nu}$, where $\nu$ is a valuation
  that assigns each $y_i$ the value $d_i$ and behaves like $w_\Data$,
  otherwise. Now supppose, for a contradiction, that $\I_{\F}$ is not
  minimal and let $\J \subsetneq \I_{\F}$ be an interpretation such
  that $\J,w_\Data \models \pathform{w_\Sigma}$. First, we show that
  the restriction $\J'$ of $\J$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$
  must coincide with $\I_{\G}$. Assuming this is not the case,
  i.e.\ $\J' \subsetneq \I_{\G}$, contradicts the minimality of
  $\I_{\G}$. Then the only possibility is that $\J(\stamp{q}{n})
  \subsetneq \I_{\F}(\stamp{q}{n})$, for some $q \in Q$. Let
  $p_1(y_1,\ldots,y_{\#(p_1)}) \arrow{a_n}{} \psi_1, \ldots,
  p_k(y_1,\ldots,y_{\#(p_k)}) \arrow{a_n}{} \psi_k$ be the set of
  transition rules in which the predicate symbol $q$ occurs. Then it
  must be the case that, for some node $p_i(d_1,\ldots,d_{\#(p_i)})$
  on the $(n-1)$-th level of $\G$, the set of children does not form a
  minimal cube from $\cube{\minsem{\stamp{\psi_i}{n}}}$, which
  contradicts the execution semantics of $\A$. \qed}

@@

The problem of checking emptiness of a given automata is undecidable,
for automata using predicates of arity two or more, even if the theory
in which the (right-hand sides of the) transition rules are written
consists of only equalities and disequalities between variables, with
no quantifiers \cite{FarzanKincaidPodelski15}. Thus, we resort to an
abstraction-refinement loop that looks for an accepted word, by firing
longer and longer sequences of input events.

Since, for a countable data domain $\Data$, the set of finite words
$\Sigma[X]^*$ over a finite set of variables $X$ is countable, the
emptiness problem is recursively enumerable in this case. As a
consequence of Lemma \ref{lemma:acceptance}, if the given automaton is
not empty, a semi-algorithm based on a systematic exploration of the
sequences of input events will eventually discover an accepted word,
provided that the first-order data theory is decidable.

>>>>

In the upcoming developments it is more convenient to work with
logical formulae defining executions. Let $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the rest of
this session. For any $i \in \nat$, we denote by $\stamp{Q}{i} =
\set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} = \set{\stamp{x}{i}
  \mid x \in X}$ the sets of time-stamped predicates and variables,
respectively. For a formula $\psi$ and $i \in \nat$, we define
$\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
formula in which all input variables and state predicates (and only
those symbols) are replaced by their time-stamped counterparts. Given
a sequence of input events $\alpha = a_1 \ldots a_n$, the \emph{path
  formula} of $\alpha$ (in $\A$) is defined as follows:
\[\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
\bigwedge_{i=1}^n \bigwedge_{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{}
  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \Longrightarrow \stamp{\psi}{i}\]
The automaton $\A$, to which $\pathform{\alpha}$ refers, will always
be clear from the context. To formalize the relation between the
low-level configuration-based execution semantics and the symbolic
path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any
execution forest $\F$ of $\A$ over $w$ is associated an interpretation
$\I_{\F}$ of the set of time-stamped predicates $\stamp{Q}{\leq n}
\isdef \set{\stamp{q}{i} \mid q \in Q, i \in [n]}$, defined as: 
\[\I_{\F}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
  q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in $\F$}}\]

@@

  ``$\supseteq$'' Let $\F$ be an execution of $\A$ over $w$. We show
  that $\I_{\F}$ is a minimal interpretation such that $\I_{\F},
  w_\Data \models \pathform{w_\Sigma}$, by induction on $n \geq
  0$. For $n=0$, $\F$ is a cube from $\cube{\minsem{\iota}}$, by
  definition. Then $\I_{\F} \models \stamp{\iota}{0}$ and moreover, it
  is a minimal such interpretation. For the inductive case $n > 0$,
  let $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\G$ be the
  restriction of $\F$ to $u$. Consequently, $\I_\G$ is the restriction
  of $\I_\F$ to $\stamp{Q}{\leq n-1}$. By the inductive hypothesis,
  $\I_{\G}$ is a minimal interpretation such that \(\I_{\G}, u_{\Data}
  \models \pathform{u_\Sigma}\). Since $\I_{\F}(\stamp{q}{n}) =
  \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid q(d_1,\ldots,d_{\#(q)})
    \text{ occurs on the $n$-th level in $\F$}}$, we have $\I_{\F},
  w_\Data \models \varphi$, for each subformula $\varphi = \forall y_1
  \ldots \forall y_{\#(q)} ~.~ \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)})
  \Rightarrow \stamp{\psi}{n}$ of $\pathform{w_\Sigma}$, by the
  execution semantics of $\A$. This is the case because the children
  of each node labeled with $q(d_1,\ldots,d_{\#(q)})$ on the
  $(n-1)$-th level of $\F$ form a cube from
  $\cube{\minsem{\psi}_\nu}$, where $\nu$ is a valuation that assigns
  each $y_i$ the value $d_i$ and behaves like $w_\Data$,
  otherwise. Now supppose, for a contradiction, that $\I_{\F}$ is not
  minimal and let $\J \subsetneq \I_{\F}$ be an interpretation such
  that $\J,w_\Data \models \pathform{w_\Sigma}$. First, we show that
  the restriction $\J'$ of $\J$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$
  must coincide with $\I_{\G}$. Assuming this is not the case,
  i.e.\ $\J' \subsetneq \I_{\G}$, contradicts the minimality of
  $\I_{\G}$. Then the only possibility is that $\J(\stamp{q}{n})
  \subsetneq \I_{\F}(\stamp{q}{n})$, for some $q \in Q$. Let
  $p_1(y_1,\ldots,y_{\#(p_1)}) \arrow{a_n}{} \psi_1, \ldots,
  p_k(y_1,\ldots,y_{\#(p_k)}) \arrow{a_n}{} \psi_k$ be the set of
  transition rules in which the predicate symbol $q$ occurs on the
  right-hand side. Then it must be the case that, for some node on the
  $(n-1)$-th level of $\G$, labeled with a configuration
  $p_i(d_1,\ldots,d_{\#(p_i)})$, the set of children does not form a
  minimal cube from $\cube{\minsem{\stamp{\psi_i}{n}}}$, which
  contradicts the execution semantics of $\A$. \qed}

@@

The problem of checking emptiness of a given automata is undecidable,
for automata using predicates of arity two or more, even only when the
theory in which the (right-hand sides of the) transition rules are
written consists of only equalities and disequalities between
variables, with no quantifiers \cite{FarzanKincaidPodelski15}.
Consequently, we cannot tackle the emptiness problem via a complete
decision procedure and resort to an abstraction-refinement loop that
tries to prove non-emptiess by discovering a word that it accepted,
firing longer and longer sequences of input events.

Assuming that the data domain is countable, the set of finite words
$\Sigma[X]^*$ is countable and the emptiness problem is recursively
enumerable. Moreover, by Lemma \ref{lemma:acceptance}, a systematic
enumeration of sequences of input events $\alpha \in \Sigma^*$ will
eventually discover an accepted word, provided that all path formulae
$\pathform{\alpha}$ belong to a decidable first-order theory.
